Model Comparisons

### 2015-11-20 models/Model_Comparison.Rmd
all model comparisons included - each testing against e6 pragmatics. One issue is that m3 tuning occurs with e11 prags - we will likely need to adjust run.rsa to accomodate this.

Set-up
------

```{r}
rm(list=ls())

library(ggplot2)
library(knitr)
library(rjson)
library(dplyr)
library(tidyr)
#setwd("~/Desktop/Projects/scalar_implicature/")
#source("analysis/useful_dplyr.R")
```

@@

# Run model functionality
```{r}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial <- function(d, alpha = 1, useprior = F, usenone = F) {
  mat <- d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
    mutate(hi = hi / sum(hi), 
           low = low / sum(low)) %>%
    select(hi, low)

  if (usenone) {
    mat$none <- c(1, 0, 0, 0, 0)
  }

  d$pred <- round(as.numeric(mapply(norm.post, d$stars, d$degree, 
                                    MoreArgs = list(m = mat, 
                                                    alpha = alpha, 
                                                    useprior = useprior))), 
                  digits=4)

@@

# run.full()
# ------------
# Run RSA with model3 (full) with alternatives
run.full <- function(d, alpha = 1, useprior = F, usenone = F) {
  mat <- d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
    mutate(hi1 = hi1 / sum(hi1),
           hi2 = hi2 / sum(hi2),
           low1 = low1 / sum(low1),
           low2 = low2 / sum(low2)) %>%
    select(hi1, hi2, low1, low2)

  if (usenone) {
    mat$none <- c(1, 0, 0, 0, 0)
  }

  d$pred <- round(as.numeric(mapply(norm.post, d$stars, d$degree, 
                                    MoreArgs = list(m = mat, 
                                                    alpha = alpha, 
                                                    useprior = useprior))), 
                  digits=4)

```

@@

## Data for Entailment models
```{r}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e8.csv")
listener = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L1_e6.csv")
# Uniform priors
priors = data.frame(stars = seq(1, 5), prior.p = rep(0.2, 5))
# Combine speaker / listener
data.partial = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

## Data for full model
```{r}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10.csv")
listener = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L1_e11.csv")
data.full = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

@@

## Run comparisons
```{r}
# Save performance output
performance.output = data.frame(model=rep(NA, 6), r2=rep(NA, 6))

# Model 1 - Entailment, no alpha tuning
# -------------------------------------
m1.noFit = data.partial %>%
  do(run.partial(., alpha = 1, useprior = T, usenone = F))
performance.output[1, ] = c("M1_noFit", round(cor(m1.noFit$listener.p, m1.noFit$pred)^2, 5))

# Model 1 - Entailment, alpha tuning
# ----------------------------------
best.alpha = tune.alpha(data.partial)
m1.fit = data.partial %>%
  do(run.partial(., alpha = best.alpha))
performance.output[2, ] = c("M1_fit", round(cor(m1.fit$listener.p, m1.fit$pred)^2, 5))

# Model 2 - Entailment, generic "None", no alpha tuning
# -----------------------------------------------------
m2.noFit = data.partial %>%
  do(run.partial(., alpha = 1, usenone = T))
performance.output[3, ] = c("M2_noFit", round(cor(m2.noFit$listener.p, m2.noFit$pred)^2, 5))

# Model 2 - Entailment, generic "None", alpha tuning
# --------------------------------------------------
best.alpha = tune.alpha(data.partial, usenone = T)
m2.fit = data.partial %>%
  do(run.partial(., alpha = best.alpha, usenone = T))
performance.output[4, ] = c("M2_fit", round(cor(m2.fit$listener.p, m2.fit$pred)^2, 5))

# Model 3 - Full, no alpha tuning
# -------------------------------
m3.noFit = data.full %>%
  do(run.full(., alpha = 1))
m3.noFit.matched = m3.noFit[matched.items, ]
performance.output[5, ] = c("M3_noFit", round(cor(m3.noFit.matched$listener.p, m3.noFit.matched$pred)^2, 5))

# Model 3 - Full, alpha tuning
# ----------------------------
best.alpha = tune.alpha(data.full, type = "full")
m3.fit = data.full %>%
  do(run.full(., alpha = best.alpha))
m3.fit.matched = m3.fit[matched.items, ]
performance.output[6, ] = c("M3_fit", round(cor(m3.fit.matched$listener.p, m3.fit.matched$pred)^2, 5))
```

Visualize
---------
```{r}
performance.output
performance.output$model = factor(performance.output$model, levels=c(
  "M1_noFit", "M1_fit", "M2_noFit", "M2_fit", "M3_noFit", "M3_fit"
))
performance.output$r2 = as.numeric(performance.output$r2)
qplot(data=performance.output, x=model, y=performance.output$r2,
      geom="bar", stat="identity",
      main = "Model performance", ylab="r^2", xlab="model") +
  geom_text(aes(label = round(r2, 3), y = r2 + 0.02), size = 3) +
  scale_y_continuous(limits=c(0, 1))

m1.noFit.G = qplot(stars, listener.p, col=degree, 
      data=m1.noFit,
      main="M1, no fit",
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m1.noFit.G

m1.fit.G = qplot(stars, listener.p, col=degree, 
      data=m1.fit,
      main="M1, no fit",
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m1.fit.G

m2.noFit.G = qplot(stars, listener.p, col=degree, 
      data=m2.noFit,
      main="M2, no fit",
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m2.noFit.G

m2.fit.G = qplot(stars, listener.p, col=degree, 
      data=m2.fit,
      main="M2, no fit",
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m2.fit.G

m3.matched.noFit.G = qplot(stars, listener.p, col=degree, 
      data=m3.noFit.matched,
      main="M3, no fit",
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.noFit.G

m3.matched.fit.G = qplot(stars, listener.p, col=degree, 
      data=m3.fit.matched,
      main="M3, no fit",
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.fit.G
```

>>>>

```{r include=FALSE}
rm(list=ls())

library(ggplot2)
library(knitr)
library(rjson)
library(dplyr)
library(tidyr)
```

@@

# Run model functionality
```{r}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
    mutate(hi = hi / sum(hi), 
           low = low / sum(low)) %>%
    select(hi, low)

  if (usenone) {
    mat$none = c(1, 0, 0, 0, 0)
  }

  d$pred = round(as.numeric(mapply(norm.post, d$stars, d$degree, 
                                    MoreArgs = list(m = mat, 
                                                    alpha = alpha, 
                                                    useprior = useprior))), 
                  digits=4)

@@

# run.full()
# ------------
# Run RSA with model3 (full) with alternatives
run.full = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
    mutate(hi1 = hi1 / sum(hi1),
           hi2 = hi2 / sum(hi2),
           low1 = low1 / sum(low1),
           low2 = low2 / sum(low2)) %>%
    select(hi1, hi2, low1, low2)

  if (usenone) {
    mat$none = c(1, 0, 0, 0, 0)
  }

  d$pred = round(as.numeric(mapply(norm.post, d$stars, d$degree, 
                                    MoreArgs = list(m = mat, 
                                                    alpha = alpha, 
                                                    useprior = useprior))), 
                  digits=4)

  return(d)
} 
```

@@

## Data for entailment models
```{r entailment_models_data, message=FALSE, warning=FALSE}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e8.csv")
listener = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L1_e6.csv")
# Uniform priors
priors = data.frame(stars = seq(1, 5), prior.p = rep(0.2, 5))
# Combine speaker / listener
data.partial = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

## Data for full model
```{r full_model_data, message=FALSE, warning=FALSE}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10.csv")
listener = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L1_e11.csv")
data.full = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

@@

## Run comparisons
```{r run_comparisons, message=FALSE}
# Save performance output
performance.output = data.frame(model=rep(NA, 6), r2=rep(NA, 6))
alphas = rep(NA, 6)

# Model 1 - Entailment, no alpha tuning
# -------------------------------------
alphas[1] = 1
m1.noFit = data.partial %>%
  do(run.partial(., alpha = alphas[1], useprior = T, usenone = F))
performance.output[1, ] = c("M1_noFit", round(cor(m1.noFit$listener.p, m1.noFit$pred)^2, 5))

# Model 1 - Entailment, alpha tuning
# ----------------------------------
alphas[2] = tune.alpha(data.partial)
m1.fit = data.partial %>%
  do(run.partial(., alpha = alphas[2]))
performance.output[2, ] = c("M1_fit", round(cor(m1.fit$listener.p, m1.fit$pred)^2, 5))

# Model 2 - Entailment, generic "None", no alpha tuning
# -----------------------------------------------------
alphas[3] = 1
m2.noFit = data.partial %>%
  do(run.partial(., alpha = alphas[3], usenone = T))
performance.output[3, ] = c("M2_noFit", round(cor(m2.noFit$listener.p, m2.noFit$pred)^2, 5))

# Model 2 - Entailment, generic "None", alpha tuning
# --------------------------------------------------
alphas[4] = tune.alpha(data.partial, usenone = T)
m2.fit = data.partial %>%
  do(run.partial(., alpha = alphas[4], usenone = T))
performance.output[4, ] = c("M2_fit", round(cor(m2.fit$listener.p, m2.fit$pred)^2, 5))

# Model 3 - Full, no alpha tuning
# -------------------------------
alphas[5] = 1
m3.noFit = data.full %>%
  do(run.full(., alpha = alphas[5]))
m3.noFit.matched = m3.noFit[matched.items, ]
m3.noFit.matched$degree = ifelse(m3.noFit.matched$degree == "hi1", "hi", "low")
m3.noFit.matched = cbind(m3.noFit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.noFit.matched$scale == data.partial$scale & m3.noFit.matched$degree == data.partial$degree)
colnames(m3.noFit.matched)[length(colnames(m3.noFit.matched))] = "e6.listener.p"
performance.output[5, ] = c("M3_noFit", round(cor(m3.noFit.matched$e6.listener.p, m3.noFit.matched$pred)^2, 5))

# Model 3 - Full, alpha tuning
# ----------------------------
alphas[6] = tune.alpha(data.full, type = "full")
m3.fit = data.full %>%
  do(run.full(., alpha = alphas[6]))
m3.fit.matched = m3.fit[matched.items, ]
m3.fit.matched$degree = ifelse(m3.fit.matched$degree == "hi1", "hi", "low")
m3.fit.matched = cbind(m3.fit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.fit.matched$scale == data.partial$scale & m3.fit.matched$degree == data.partial$degree)
colnames(m3.fit.matched)[length(colnames(m3.fit.matched))] = "e6.listener.p"
performance.output[6, ] = c("M3_fit", round(cor(m3.fit.matched$e6.listener.p, m3.fit.matched$pred)^2, 5))
```

# Plots

Overall model performance: $r^2$
```{r overall_perf}
performance.output
performance.output$model = factor(performance.output$model, levels=c(
  "M1_noFit", "M1_fit", "M2_noFit", "M2_fit", "M3_noFit", "M3_fit"
))
performance.output$r2 = as.numeric(performance.output$r2)
qplot(data=performance.output, x=model, y=performance.output$r2,
      geom="bar", stat="identity",
      main = "Model performance", ylab="r^2", xlab="model") +
  geom_text(aes(label = round(r2, 3), y = r2 + 0.02), size = 3) +
  scale_y_continuous(limits=c(0, 1))
```

## Individual model performance
```{r individual_plots}
m1.noFit.G = qplot(stars, listener.p, col=degree, 
      data=m1.noFit,
      main=paste("M1, no fit\nalpha: ", alphas[1]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m1.noFit.G

m1.fit.G = qplot(stars, listener.p, col=degree, 
      data=m1.fit,
      main=paste("M1, fit\nalpha: ", alphas[2]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m1.fit.G

m2.noFit.G = qplot(stars, listener.p, col=degree, 
      data=m2.noFit,
      main=paste("M2, no fit\nalpha: ", alphas[3]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m2.noFit.G

m2.fit.G = qplot(stars, listener.p, col=degree, 
      data=m2.fit,
      main=paste("M2, fit\nalpha: ", alphas[4]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m2.fit.G

m3.matched.noFit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m3.noFit.matched,
      main=paste("M3, no fit\nalpha: ", alphas[5]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.noFit.G

m3.matched.fit.G = qplot(stars, e6.listener.p, col=degree, 
      data=m3.fit.matched,
      main=paste("M3, fit\nalpha: ", alphas[6]),
      ylab="Posterior p(rating | word)") + 
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
m3.matched.fit.G

m3.matched.fit.CorPlot = ggplot(data=m3.fit.matched, aes(y=pred, x=e6.listener.p)) +
  geom_point(aes(colour = scale)) +
  geom_smooth(method=lm) +
  ggtitle("Model performance: M3 fit") +
  labs(x = "human judgments", y="model prediction")

m3.matched.fit.CorPlot
```

# Other exploration
```{r}
bad.predictions = m3.fit.matched$pred > 0 &
                  m3.fit.matched$e6.listener.p == 0 &
                  (m3.fit.matched$pred - m3.fit.matched$e6.listener.p > 0.05)
```
